The unphysicality of Hilbert spaces

### 2023-06-20 Articles/2023-HilbertQuantum/HilbertQuantum.tex
Updated introduction

Von Neumann's \textit{Mathematische Grundlagen der Quantenmechanik} (\citeyear{von_neumann_mathematische_1996}) crowned Hilbert spaces as (i) the unifying framework for the states in Schrödinger's wave mechanics and Heisenberg-Born-Jordan's matrix mechanics that is (ii) mathematically rigorous. Arthur Wightman, one of the people who followed von Neumann's aims to unify and axiomatize quantum field theory (\citeyear{wightman_quantum_1956}), took his book to be the fulfillment of a project Hilbert himself kicked off with his problems on the foundations of mathematics (\citeyear[p. 157]{browder_hilberts_1976}): to axiomatize the physical sciences. Despite the undeniable importance of wanting to set physics on a firm mathematical footing, we must remember that by ``axioms,'' they meant ``physically motivated constraints'' more than statements that can be immune to question, rethink, and, if necessary, revise.\footnote{For a more extended discussion of von Neumann's axiomatics and their physical inputs, see \citep{redei_soft_2006}.}\\

In this letter, we focus on an assumption that poses an array of interesting questions: that quantum states are vectors in Hilbert spaces. At first glance, some of their shortcomings (states are rays, not vectors; the Dirac delta is a distribution, not a function; some elements from $L^2(\mathbb{R})$ don't seem apt candidates to be wave functions) are well-known, but deemed unproblematic. Additionally, a handful of theorems (Gleason, Riesz, Wigner, Stone, Stone-von Neumann, Wigner, etc.) rely heavily on the structure of a Hilbert space. While these results are rarely mentioned when students are first introduced to quantum mechanics, they are crucial to substantiate the claim that Hilbert spaces are the right structure to characterize quantum states. Therefore, it's not a surprise that Hilbert spaces have become fossilized as parts of axiomatic systems\footnote{Algebraic quantum field theory (see, for instance, \citep{haag_local_1996}) inverts the narrative of ``quantum states are normed vectors in a Hilbert space, and observables are bounded, self-adjoint operators acting on that space'' to ``canonical (anti)commutation relations of observables generate an algebra, and states are assignments of expectation values to elements of the algebra.'' As such, it doesn't \textit{need} to talk about Hilbert spaces to do quantum mechanics, but it has resources, like the GNS construction, that allow their advocates to say that, since they're recovering Hilbert spaces from algebras of observables, what they're doing ends up being similar to the quantum mechanics that's more familiar to physicists.} without even considering whether we can do quantum mechanics with weaker assumptions. So, despite all their success, perhaps it's time to ask if using Hilbert spaces to describe quantum states, an assumption whose physical motivation was meant to be revised as time passed, has become a dogma.

>>>>

Hilbert spaces have been the cornerstone of quantum mechanics since von Neumann (\cite{von_neumann_mathematische_1996}). Their use made available a vast number of theorems (Gleason, Riesz, Wigner, Stone, etc.) that made up the mathematical backbone of the theory and also represented a role model of mathematical rigor that people working on quantum field theory would try to achieve decades later \cite{haag_discussion_2010}. Wightman even took von Neumann's book to be the fulfillment of a project Hilbert himself kicked off with his problems on the foundations of mathematics (\cite[p. 157]{browder_hilberts_1976}): to axiomatize the physical sciences. But, despite the undeniable importance of wanting to set physics on a firm mathematical footing, we must remember that by ``axioms,'' they meant ``physically motivated constraints'' more than statements that can be immune to question, rethink, and, if necessary, revise.\footnote{For a more extended discussion of von Neumann's axiomatics and their physical inputs, see \cite{redei_soft_2006}.}

Now, there are well-known shortcomings of Hilbert spaces, but they are deemed unproblematic: states are rays, not vectors; the Dirac delta is a distribution, not a function; and some elements from $L^2(\mathbb{R})$ don't seem apt candidates to be wave functions. Given the importance Hilbert spaces have had and the issues we have let slide, it's not a surprise that they have become fossilized as parts of axiomatic systems\footnote{Even axiomatic formulations of quantum field theory (see e. g. \cite{haag_local_1996}) that don't need to use Hilbert spaces to make meaningful claims about quantum-mechanical observables and states can reassure physicists unfamiliar with this approach by recovering the Hilbert spaces they know and love through the GNS construction.} without even considering whether we can do quantum mechanics with weaker assumptions. So, despite all their success, perhaps it's time to ask if using Hilbert spaces to describe quantum states, an assumption whose physical motivation was meant to be revised as time passed, has become a dogma.
